Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-03-08 TEBenchmarkPaper/article.tex
Fleshed out application section

To solve both of these problems we instead \emph{sample} a subset of
definitions. Given a sample size, we choose a subset of that many definitions,
and provide those definitions as our MTE input. We generate a corresponding
ground truth by selecting those theorems from the corpus which only contain
references to the sampled definitions.

Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
given sample size, increasing the size of the corpus (which, as detailed above,
is advantageous) makes it less likely that a chosen sample will contain all of
the definitions required by a theorem. The majority of such samples would hence
have an empty ground truth, and thus $0$ precision and undefined recall
\emph{independent} of what the MTE system produces! This is clearly undesirable
as an evaluation method.

Instead, we only choose samples which admit at least one of the corpus theorems.
We could do this using rejection sampling, but it is more efficient to sample
the \emph{theorem statements}, weighted in proportion to the number of
definitions they reference. Once a theorem is chosen, we use the referenced
definitions as our sample, padded up to the required size by uniformly sampling
from the remaining definitions (we ignore theorems which reference more
definitions than our sample size).

When we form the ground truth for such samples, we're guaranteed to find at
least one applicable theorem from the corpus (the one we chose as our starting
point), and hence have a meaningful precision and recall.

@@

    \node (in)  [io]{Theorem Proving Benchmark (TIP)};
    \node (sep) [process,       below=1cm of in         ]{Separate definitions from theorems};
    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};

@@

We have applied our methodology to evaluating QuickSpec, %TODO:cite
a library for conjecturing equations about functions written in the Haskell
programming language. We use the Tons of Inductive Problems (TIP) theorem
proving benchmark to generate our corpus.

We use QuickSpec version 0.9.6 to demonstrate our benchmarking methodology. This
is a Haskell library which takes a \emph{signature} containing Haskell functions
and (universally quantified) typed variables, and outputs conjectured equations
involving these terms.

The QuickSpec algorithm first enumerates all well-typed expressions up to a
certain depth (3 by default), involving the given terms. Expressions of the same
type are assumed to be equivalent, and grouped into equivalence classes. A
testing process then tries to disprove this assumption by instantiating each
variable with a randomly chosen value (using the QuickCheck testing library),
evaluating the expressions and comparing the results. The equivalence classes
are split to separate any expressions which are observed to be different, and
the process repeats with new random values.

After 500 rounds of testing, any expressions still sharing the same equivalence
class are conjectured to be equal. Finally, a congruence closure algorithm is
applied to these equations, to find a minimal set.

\begin{itemize}
\item We must decide what variables to include. We add three for each type that
  appears as a function argument, except for types which have no QuickCheck data
  generator.
\item We must \emph{monomorphise} all types. For example, functions like
  \texttt{cons} are \emph{polymorphic}: they build lists of any element type.
  We need to pick a specific type in order to generate random values, so we
  resolve this (arbitrarily) by picking
  \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
    \texttt{*}, and \texttt{[]} (Haskell's list type constructor) for those of
    kind \texttt{* -> *}. If these violate some constraint, we pick a suitable
    type non-deterministically from the module scope, or skip these variables
    completely if none is found.}
\item Haskell functions only take one argument, as as in the $\lambda$-calculus,
  so functions of multiple arguments are simulated by currying.
  Unfortunately QuickSpec cannot compare functions for equality, so we must give
  each function in our signature an explicit arity. Arities from zero to five
  are supported, so we avoid partial application by picking the highest that is
  type-correct.
\end{itemize}

To find out which functions to include in a signature, we use a plugin for the
GHC compiler which tells us all of the functions defined in a package. We
include all of those which type-check when passed to QuickSpec.

@@

\begin{sloppypar}
  We rewrite the TIP theorems (our ground truth) to reference these expanded
  forms instead of the raw constructors and destructors, and include these
  functions in our samples in lieu of the raw expressions. Note that these
  destructor wrappers are \emph{partial} functions (e.g.
  \texttt{destructor-head} and \texttt{destructor-tail} are undefined for the
  input \texttt{Nil}), which complicates their translation to proof assistants
  like Isabelle.
\end{sloppypar}

Another complication is TIP's ``native'' support for booleans and integers,
which allows numerals and symbols like \texttt{+} to appear without any
accompanying definition. To ensure consistency in the translations, we replace
all occurrences of such expressions with standard definitions written with the
``user-level'' \texttt{declare-datatypes} and \texttt{define-fun} mechanisms.

When we add all of these generated types and functions to those in TIP, we get a
total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
and we choose to only sample from those 182 which occur in at least one theorem
statement (this removes ambiguity about which \emph{definitions} count as
``interesting'' and which are just ``implementation details'' for other
definitions).

TIP includes a tool to translate its definitions into Haskell and Isabelle,
including suitable comparison functions and random data generators. We translate
all 269 unique definitions into a single module/theory which is imported on each
run of the MTE tools (although only those functions which appear in the current
sample are actually explored).

\subsection{Results}

\begin{figure}
  \centering
  \input{time.pgf}
  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
    is an individual run, with 30 per size except for 2 (29). Colour indicates
    the total number of conjectures produced by each run, with failing processes
    shown in red. Some points are shifted horizontally to prevent overlaps.
    Standard box plots are included, with whiskers showing points at
    1.5~$\times$~the inter-quartile range.}
  \label{figure:quickspec_runtimes}
\end{figure}

\begin{figure}
  \centering
  \input{prec.pgf}
  \caption{Precision and recall of all runs (spread horizontally to avoid
    overlaps). Point colour shows how many conjectures were in the ground truth
    set for each run. Lines show the mean value (average of ratios) and the
    shaded region is standard deviation, assuming a binomial distribution.}
  \label{figure:quickspec_precRec}
\end{figure}

We applied our sampling methodology to the 182 definitions which appear in TIP
theorems, for even sample sizes up to 20. We chose 30 samples for each size,
although one sample of size 2 was discarded as it was a duplicate. A timeout of
5 minutes was imposed, since preliminary experiments showed most QuickSpec runs
to either finish before this time, or still be running after an hour. The
running times are shown in figure~\ref{figure:quickspec_runtimes}.

The only errors we encountered were timeouts (shown in red), which occurred for
all sample sizes but are more common for larger samples (over half the runs for
sample size 20 timed out). Runs which did not time out mostly finished within 30
seconds, generating few or no conjectures; those taking longer generated more
conjectures, with the most being 116 conjectures from a sample of size 16.

Equations generated by QuickSpec and those from the ground truth were normalised
by numbering their variables and putting the left- and right-hand-sides in
lexicographical order. These two sets were then compared syntactically for
precision and recall, with the results shown in
figure~\ref{figure:quickspec_precRec}.

We use a binomial distribution to model precision and recall, as if each
conjecture is drawn from a pool of wanted/unwanted results (in the case of
precision) or found/missed ground truths (in the case of recall), with the mean
of these distributions estimating the proportions of each pool's split.

There is a clear negative trend in precision as the sample size increases,
indicating that a more aggressive post-processing filter may be useful as we
explore larger theories. Some small samples reached full precision, but nothing
above size 6 exceeded 50\%.

Recall shows a slight downward trend but is mostly flat, averaging between 20\%
and 50\%. This is expected, since adding more definitions to a QuickSpec
signature will never reduce the conjectures it will find. Those conjectures
which QuickSpec failed to find include terms which QuickSpec cannot synthesise
(such as conditional equations, inequalities, terms larger than the search depth
and those with anonymous functions) as well as equations which are removed as
``uninteresting'' by the final congruence closure step.

>>>>

To solve both of these problems we instead \emph{sample} a subset of
definitions. Given a sample size, we choose a subset of that many definitions,
and provide those definitions as input to our MTE tool. We generate a
corresponding ground truth by selecting those theorems from the corpus which
``depend on'' (contain references to) \emph{only} the definitions in that
sample. Transitive dependencies aren't required (e.g. a theorem which only
references \texttt{times} does not depend on \texttt{plus}, even if
\texttt{plus} occurs in the definition of \texttt{times}).

Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
given sample size, increasing the size of the corpus (which, as detailed above,
is advantageous) makes it less likely that a chosen sample will contain all of
a theorem's dependencies. The majority of such samples would hence have an empty
ground truth, and thus $0$ precision and undefined recall \emph{independent} of
what the MTE tool produces! This is clearly undesirable as an evaluation
method.

Instead, we only allow a sample if it contains all dependencies of at least one
theorem. We could do this using rejection sampling, but it is more efficient to
pick a theorem, weighted in proportion to their number of dependencies (ignoring
those with too many dependencies to fit in our sample). We use those
dependencies as our sample, padded up to the required size with definitions
chosen uniformly from the remainder.

The ground truth for such samples is guaranteed to contain at least one theorem
(the one we picked), and hence the precision and recall will depend meaningfully
on the tool's output.

@@

    \node (in)  [io]{Theorem Proving Benchmark};
    \node (sep) [process,       below=1cm of in         ]{Separate definitions from theorems};
    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};

@@

We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
benchmark as our ground truth corpus. TIP 0.2 contains 343 theorem statements,
which depend on 182 distinct function definitions.

To determine our benchmarking parameters we ran some initial tests on both tools
for samples sized between 1 and 100, for an hour each. Most QuickSpec runs
either finished within 200 seconds or not at all, and sizes above 20 mostly
timed out. IsaCoSy mostly finished within 300 seconds on sample sizes up to 4,
but by size 8 was mostly timing out; its few successes above this took thousands
of seconds each, which we deemed infeasibly long.

We decided to benchmark sample sizes up to 20, since neither tool seemed to
perform well beyond this, and we followed the statistical ``rule of thumb'' of
picking 30 samples for each size\footnote{Only 29 samples of size 2 were
  analysed since one was a duplicate, due to the constraint that samples contain
  at least one theorem's dependencies.}. This gave $20$ sizes $\times 30$
samples $\times 2$ tools $= 1200$ runs. To keep the total time down to a few
days we chose a timeout of 300 seconds, since that covered most of the
successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
diminishing returns.

We normalised the conjectures generated by each tool, as well as the theorems
from the ground truth, by numbering their free variables from left to right and
using de Bruijn indices for bound variables. Since both tools can only output
equations, we also sorted their left- and right-hand-sides into lexicographical
order. Precision and recall were calculated by syntactically comparing the sets
of generated conjectures to those of their ground truth. Other than equation
symmetry, we specifically \emph{ignored} logical relationships between
syntactically distinct statements, such as one equation being implied by
another. The main reason is to avoid second-guessing the ground truth: the
theorems' authors chose to state them that way, and it is not our place to
decide whether an alternative form is ``just as good'' (for example, a more
general statement is more widely applicable, but it might also be harder to
understand).

@@

\begin{sloppypar}
  We rewrite the TIP theorems (our ground truth) to reference these expanded
  forms instead of the raw constructors and destructors, and use these functions
  in our samples in lieu of the raw expressions. Note that these destructor
  wrappers are \emph{partial} functions (e.g. \texttt{destructor-head} and
  \texttt{destructor-tail} are undefined for the input \texttt{Nil}), which
  complicates their translation to proof assistants like Isabelle.
\end{sloppypar}

Another complication is TIP's ``native'' support for booleans and integers,
which allows numerals and symbols like \texttt{+} to appear without any
accompanying definition. To ensure consistency in the translations, we replace
all occurrences of such expressions with standard definitions written with the
``user-level'' \texttt{declare-datatypes} and \texttt{define-fun}
mechanisms.~\footnote{\texttt{Boolean} has \texttt{true} and \texttt{false}
  constructors; \texttt{Natural} has \texttt{zero} and \texttt{successor};
  \texttt{Integer} has unary \texttt{positive} and \texttt{negative}
  constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
  symmetery.}

When we add all of these generated types and functions to those in TIP, we get a
total of 3598 definitions. Removing $\alpha$-equivalent duplicates leaves 269,
and we choose to only sample from those 182 functions which occur in at least
one theorem statement (this removes ambiguity about which \emph{definitions}
count as ``interesting'' and which are just ``implementation details'' for other
definitions).

TIP comes with software to translate its definitions into Haskell and Isabelle
code, including comparison functions and random data generators suitable for
QuickSpec. We translate all 269 unique definitions into a single module/theory
which is imported on each run of the tools, although only those functions which
appear in the current sample are actually explored.

We benchmarked QuickSpec version 0.9.6, which is tool written in Haskell for
conjecturing equations involving a given set of Haskell functions and
universally-quantified variables (a \emph{signature}).

The QuickSpec algorithm first enumerates all well-typed expressions containing
terms from the signature, up to a chosen depth (3 by default). Expressions of
the same type are assumed to be equal, and grouped into equivalence classes. A
testing process then tries to refute this assumption by instantiating each
variable with a randomly chosen value (using the QuickCheck testing library),
evaluating the expressions with those values and comparing the results. The
equivalence classes are split up to separate any expressions which are observed
to be different, and the process is repeated with new random values.

After 500 rounds of testing, any expressions still sharing the same equivalence
class are conjectured to be equal. Finally, a congruence closure algorithm is
applied to these equations, to remove redundancies and special-cases.

\begin{itemize}
\item We must decide what variables to include. We choose to add three variables
  for each type that appears as a function argument, except for types which have
  no QuickCheck data generators.
\item We must \emph{monomorphise} all types. For example, functions like
  \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
  element type, but we need to pick a specific type in order to know which
  random value generator to use. We resolve this (arbitrarily) by picking
  \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
    \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
    \texttt{[]} (Haskell's list type constructor). If these violate some
    type class constraint, we pick a suitable type non-deterministically from
    those in scope during compilation; if no suitable type is found, we give up
    and don't include the  variable.}
\item Haskell functions are ``black boxes'', which QuickSpec can't compare
  during its exploration process. They are also curried, always taking one
  argument but potentially returning another function. QuickSpec lets us assign
  an arity to each function in the signature, from 0 to 5, so we pick the
  highest that is type-correct, since this avoids a proliferation of
  incomparable, partially-applied functions.
\end{itemize}

\begin{figure}
  \centering
  \input{quickspectime.pgf}
  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
    is an individual run, with 30 per size except for 2 (29). Colour indicates
    the total number of conjectures produced by each run, with failing processes
    shown in red. Points are spread out horizontally to prevent overlaps.
    Box plots show median and quartiles, with whiskers showing points at
    1.5$\times$~inter-quartile range.}
  \label{figure:quickspec_runtimes}
\end{figure}

\begin{figure}
  \centering
  \input{quickspecprec.pgf}
  \caption{Precision and recall of successful QuickSpec runs (spread
    horizontally to avoid overlaps). Point colour shows how many conjectures
    were in the ground truth set for each run. Lines show the combined
    proportion for each sample size (ratio of averages) and the shaded region is
    the sample standard deviation.}
  \label{figure:quickspec_precRec}
\end{figure}

The time taken to explore samples of different sizes is shown in
\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red, and
these occurred for all sample sizes but are more common for larger samples (over
half the runs for sample size 20 timed out). Runs which succeeded mostly
finished within 30 seconds, generating few or no conjectures; those taking
longer generated more conjectures, with the most being 116 conjectures from a
sample of size 16.

We model precision and recall as simple Bernoulli processes: the generated
conjectures (wanted or not) are like coin tosses whose bias is the precision;
likewise, the ground truth theorems (found or not) are like coin tosses biased
by the recall.

QuickSpec's precision gets worse as the sample size grows. Since QuickSpec
generates monotonically more conjectures as definitions are added to a signature
(assuming sufficient testing), it can't be finding fewer correct conjectures at
larger sizes. This is supported by the relative flatness of the recall results.
Rather, the number of conjectures generated is increasing at a higher rate than
the size of the ground truth. These extra conjectures may involve the
``padding'' definitions in a sample which don't contribute to its ground truth,
or may be ``uninteresting'' relationships between the dependencies of different
ground truth theorems.

This indicates two potential improvements to the QuickSpec algorithm (as far as
this benchmark is concerned). The deluge of generated conjectures could be
filtered by another post-processing step down to a more desirable sub-set.
Alternatively, large signatures could be filtered to remove ``padding''
definitions (those whose conjectures are unlikely to be ``interesting'') and
divided up into multiple smaller signatures to separate those definitions whose
relationships are ``uninteresting''. Both of these methods could improve the
precision, although care would be needed to avoid a large reduction in recall.
The latter option could also improve the running time, since multiple smaller
signatures may be faster to explore than a single large one.

QuickSpec's recall is limited by two factors: the algorithm is unable to
synthesise some theorem statements, such as conditional equations, inequalities,
terms larger than the search depth and those containing anonymous functions.
The congruence closure algorithm used as a post-processor may also be removing
``interesting'' results, for example if we found an ``uninteresting'' result
which is more general.

\subsection{IsaCoSy}

IsaCoSy is a tool for the Isabelle proof assistant, mostly written in Standard
ML. We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
with the 2015 version of Isabelle.

The following issues had to be overcome for the benchmarking to be successful:

\begin{itemize}
  polyrec can't be represented
  constructors were being included
  proofs had to be skipped
  partial functions needed adding to the constraints
\end{itemize}

\begin{figure}
  \centering
  \input{isacosytime.pgf}
  \caption{Running times of IsaCoSY on theories sampled from TIP. Each point
    is an individual run, with 30 per size except for 2 (29). Colour indicates
    the total number of conjectures produced by each run, with failing processes
    shown in red (timeouts and out of memory). Points are spread out
    horizontally to prevent overlaps. Box plots show median and quartiles, with
    whiskers showing points at 1.5$\times$~inter-quartile range.}
  \label{figure:isacosy_runtimes}
\end{figure}

\begin{figure}
  \centering
  \input{isacosyprec.pgf}
  \caption{Precision and recall of all successful IsaCoSy runs (spread
    horizontally to avoid overlaps). Lines show the combined proportion for each
    sample size (ratio of averages) and the shaded region is the sample standard
    deviation.}
  \label{figure:isacosy_precRec}
\end{figure}

The most striking result is how rapidly IsaCoSy's running time increases with
sample size: from size 8 upwards, all runs failed (most timed out, a few aborted
early due to running out of memory).

We compare the performance of QuickSpec and IsaCoSy using a similar protocol to
the Speedup-Test of Touati, Worms and Briais~\cite{touati2013speedup}. We differ
by timing a wider variety of inputs: running many samples once, rather than a
few samples many times and taking the medians. Since each sample is run through
both tools, we can also gain statistical power by using the \emph{paired} form
of the Wilcoxon signed-rank test.
